Dependent Object Types

### 2012-09-14 doc/fool/article.tex
new abs and rev intro

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of refinement
types. It makes no attempt to model inheritance or mixin
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

We show that DOT and its patched-up variants are not syntactically
sound, by exhibiting counterexamples to preservation. Nevertheless, we
sketch a proof of type-safety of the calculus via step-indexed logical
relations.
\end{abstract}

@@

In the present calculus, we replace Scala's compound types by
classical intersection types, which are commutative. We also
complement this by classical union types. Intersections and unions
form a lattice wrt subtyping. This addresses another problematic
feature of Scala: In Scala's current type system, least upper bounds
and greatest lower bounds do not always exist. Here is an example:
given two traits
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
The greatest lower bound of \code@A@ and \code@B@ is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B {
    type T < ...
  }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala.

This is problematic because glbs and lubs play a central role in
Scala's type inference. The absence of universal glbs and lubs makes
type inference more brittle and more unpredictable.

@@

We propose DOT as a core calculus for path-dependent types. We present
the calculus formally in section \ref{dot-calculus} and through examples in
section~\ref{dot-examples}. Though we show that the calculus does not
satisfy the standard theorem of preservation in
section~\ref{dot-preservation}, we contribute a proof sketch of type
safety using logical relations in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

>>>>

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types, abstract type members and its mixture of nominal
and structural typing through the use of refinement types.

DOT does not precisely describe what's currently in Scala: it is more
normative than descriptive, since our goal is to explore
simplifications and unifications of concepts in the type system. In
particular, DOT unifies Scala's constructs for type members, and
provides classical intersection and union types, which simplify
Scala's compound types and glb/lub computations.

In this paper, we present and motivate the DOT calculus, our core
calculus for path-dependent types, both formally and informally. We
also discuss our work-in-progress to prove type-safety of the calculus.
\end{abstract}

@@

In the present calculus, we replace Scala's compound types by
classical intersection types, which are commutative. We also
complement this by classical union types. Intersections and unions
form a lattice wrt subtyping. This addresses another problematic
feature of Scala: In Scala's current type system, least upper bounds
and greatest lower bounds do not always exist. Here is an example:
given two traits \code{A} and \code{B}, where each declares an
abstract upper-bounded type member \code{T},
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
the greatest lower bound of \code{A} and \code{B} is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B {
    type T < ...
  }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala.

This is problematic because glbs and lubs play a central role in
Scala's type inference. For example, in to infer the type \code{if} expression such as
\begin{lstlisting}
  if (cond) ((a: A) => c: C) else ((b: B) => d: D)
\end{lstlisting}
type inference tries to compute the glb of \code{A} and \code{B} and
the lub of \code{C} and \code{D} The absence of universal glbs and
lubs makes type inference more brittle and more unpredictable.

@@

We propose DOT as a core calculus for path-dependent types. We first
motivate the calculus informally in section~\ref{dot-motiv}. We
present the calculus formally in section \ref{dot-calculus} and
through examples in section~\ref{dot-examples}. Though we show that
the calculus does not satisfy the standard theorem of preservation in
section~\ref{dot-preservation}, we contribute a proof sketch of type
safety using logical relations in section~\ref{dot-type-safety}. In
section~\ref{discussion}, we discuss choices and variants of the
calculus, as well as related work, and conclude in
section~\ref{conclusion}.

\section{Motivation}\label{dot-motiv}
%TODO
